Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers

### 2025-01-28 endofquarterpaper.qmd
minor edits

One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete. Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis are used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.

Similar analyses performed upon bead work patterns from the Aurignacian technocomplex as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d]. Specifically, the groups are identified using seriation, correspondence and GIS analyses. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. As said previously, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.

In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects. Firstly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance. We then identify clusters using seriation, network, and Principal Coordinate analyses. Finally, we test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.

@@

Mobility for various reasons is commonly found in hunter-gatherer groups. Therefore, one explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility. The mobility may have occurred for all the above reasons. One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.

>>>>

A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019]. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian period. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016]. In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

The purpose of this study is to contribute to our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic. Previous works have performed cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account. For example, multivariate analyses of Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence]. The data set consisted of 134 discrete beads types from both burial and occupation sites, as distinguished by shape and material. Seriation analysis, PCoA, and network analysis were used to identify the groups. The geographically discrete clusters identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, is tested using a Mantel test, determining that geographic distance cannot solely account for the bead distribution.

Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity. The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d]. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. Availability of raw materials was considered ~in regards to~ the potential explanation that observed patterns are determined by ~such~. ~It was concluded that, due to observed patterns in the data that countered that explanation~, [BM: They found that] the differences in personal ornamentation could not be explained by material availability.

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects. Firstly, we identify clusters using seriation analysis, network analysis, and PCoA. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.

@@

One explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility. @Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Similarly, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity. One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.
